Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-28 paper.tex
Paragraph on inference methods.

There seems to be little point in continuing the huge duplication of effort we
have seen. It is not possible to achieve the performance levels of \msprime\
without investing substantially in software engineering. As demonstrated here,
\msprime\ is extensible and we would argue that developing this simulator as an
open-source community asset is of more benefit to all. Implementing a
coalescent simulator may once have been a good way for (e.g.) a student to
learn about programming and simulation, but it is no longer reasonable to
expect this. There is simply too much engineering required. For example, as of
the 1.0.0 release \msprime\ consisted of around 13K lines of C and 11K
lines of Python, with test suites of 122 C test cases (7K lines of code) and
1350 Python test cases (22K lines of code), along with a suite of 217
statistical validation tests (6K lines of code). It requires a sustained effort
over years to produce well engineered, high-quality software. It may reasonably
be asked if students would learn more from contributing to an existing project
with established high-quality engineering standards, or from producing their own
software in a self-taught manner.

A significant advantage to simulating the coalescent with recombination
backwards in time using Hudson's algorithm rather than
left-to-right
approaches~\citep{wiuf1999recombination,chen2009fast,staab2015scrm}
is the ease with which additional processes can be integrated.
We have seen that processes such as gene conversion, complex
demography, instantaneous bottlenecks, multiple merger caolescents
and selective sweeps have been integrated seamlessly into
the simulations. Incorporating any of these extensions into
either the exact or approximate left-to-right processes would require
complex derivations. Because we are going backwards
in time, processes can be modelled as competing exponentials
using Gillespie's algorithm \jkcomment{I think?},
and therefore in principle we can
add as many extra processes as we like.
\jkcomment{so, backwards in time
sims in msprime is extensible for the future.}

There is also the welcome development of libraries such as
Quetzal~\citep{becheler2019quetzal} and libsequence~\citep{thornton2014cpp}
which break this pattern of monolithic simulation software.

% Finish up with a paragraph saying that there loads of exciting
% new things that simulation can do. We need MORE sims!!!
The ability to infer trees at scale for the first time
in the presence of
recombination~\citep{harris2019database,kelleher2019inferring,
speidel2019method,tang2019genealogy}
% also SARGE
opens the possibility for new applications of coalescent simulations.
With lots of applications, e.g., [cite Osmond and Coop].
Also \stdpopsim\ is cool~\citep{adrion2020community}

>>>>

% There seems to be little point in continuing the huge duplication of effort we
% have seen. It is not possible to achieve the performance levels of \msprime\
% without investing substantially in software engineering. As demonstrated here,
% \msprime\ is extensible and we would argue that developing this simulator as an
% open-source community asset is of more benefit to all. Implementing a
% coalescent simulator may once have been a good way for (e.g.) a student to
% learn about programming and simulation, but it is no longer reasonable to
% expect this. There is simply too much engineering required. For example, as of
% the 1.0.0 release \msprime\ consisted of around 13K lines of C and 11K
% lines of Python, with test suites of 122 C test cases (7K lines of code) and
% 1350 Python test cases (22K lines of code), along with a suite of 217
% statistical validation tests (6K lines of code). It requires a sustained effort
% over years to produce well engineered, high-quality software. It may reasonably
% be asked if students would learn more from contributing to an existing project
% with established high-quality engineering standards, or from producing their own
% software in a self-taught manner.

% A significant advantage to simulating the coalescent with recombination
% backwards in time using Hudson's algorithm rather than
% left-to-right
% approaches~\citep{wiuf1999recombination,chen2009fast,staab2015scrm}
% is the ease with which additional processes can be integrated.
% We have seen that processes such as gene conversion, complex
% demography, instantaneous bottlenecks, multiple merger caolescents
% and selective sweeps have been integrated seamlessly into
% the simulations. Incorporating any of these extensions into
% either the exact or approximate left-to-right processes would require
% complex derivations. Because we are going backwards
% in time, processes can be modelled as competing exponentials
% using Gillespie's algorithm \jkcomment{I think?},
% and therefore in principle we can
% add as many extra processes as we like.
% \jkcomment{so, backwards in time
% sims in msprime is extensible for the future.}

Recent breakthroughs in methods to infer genetic ancestry in recombining
organisms~\citep{kelleher2019inferring,
speidel2019method,wohns2021unified,schaefer2021ancestral,speidel2021inferring}
have made it possible to estimate such ancestry at scale for
the first time~\citep{harris2019database,tang2019genealogy}.
Given such inferred ancestry, many exciting applications
become possible. For example, \cite{osmond2021estimating}
developed a method to estimate the location of genetic ancestors
based on inferred trees, and other uses are sure to follow.
Since the inferred genetic ancestry
becomes the input for other downstream inferences, it is vitally
important that these primary inferences are thoroughly validated,
with the detailed properties of the inferred ancestries
catalogued and understood.
As we have seen, \msprime\---through the succinct tree sequence
data structure and \tskit\ library---can provide simulated
genetic ancestry at any scale or degree of detail required, and should
provide a valuable source of ground-truth for evaluating
these methods.
